---
title: "R output for A2"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Question 2. a)
```{r q2}
y = c(5.5,5.9,6.5,5.9,8,9,10,10.8)
X = matrix(c(rep(1,8),7.2,10,9,5.5,9,9.8,14.5,8,8.7,9.4,
             10,9,12,11,12,13.7,5.5,4.4,4,7,5,6.2,5.8,3.9),8,4)
b = solve(t(X)%*%X,t(X)%*%y)
e = y - X%*%b
n = dim(X)[1]
p = dim(X)[2]
s2 = sum(e^2)/(n-p)

b
s2
```

Question 2. b)
```{r q2b}
c = solve(t(X)%*%X)
c
```

Question 2. c)
```{r q2c}
s = sqrt(sum(e^2)/(n-p))
alpha = 0.01
ta = qt(1-alpha/2, df=(n-p))
t = c(1,8,9,5)
ttb = t(t)%*%b
CI = c(ttb) + c(-1,1)*c(ta*s*sqrt(t(t)%*%solve(t(X)%*%X)%*%t))

CI
```

Question 2. d)
```{r q2d}
for (alpha in seq(0.01, 0.15, by = 0.0005)) {
  # t_alpha given an alpha value
  ta = qt(1-alpha/2, df=(n-p)) 
  # Generate Prediction Interval given alpha
  PI = c(ttb) + c(-1,1)*c(ta*s*sqrt(1+t(t)%*%solve(t(X)%*%X)%*%t))  
  if (round(PI[1],3) == 4.012 && round(PI[2],3) == 7.087) {
    print(alpha)
    print(round(PI,3))
  }
}
```

Question 2. e)
```{r q2e}
SSRes = t(y-X%*%b)%*%(y-X%*%b)
CorrectedSSReg = t(y)%*%X%*%b - sum(y)^2/n
k = 3 # num parameters
Fstat = (CorrectedSSReg/k)/(SSRes/(n-k-1))
Fval = qf(0.95,k,n-k-1)

Fstat
Fval
Fstat > Fval
```

Question 4. a)
```{r q4a}
data(state)
statedata <- data.frame(state.x77, row.names=state.abb, check.names=TRUE)

pairs(statedata)

statedata$Area = log(statedata$Area) # log Area
statedata$Illiteracy = log(statedata$Illiteracy) # log Illiteracy
```

Question 4. b)
```{r q4b}
basemodel = lm(Murder ~ 1, data=statedata)

add1(basemodel, scope= ~ . + Population + Income + Illiteracy + Life.Exp + HS.Grad + Frost + Area, data=statedata, test="F")
# we add Life.Exp
model1 = lm(Murder ~ Life.Exp, data=statedata)
add1(model1, scope= ~ . + Population + Income + Illiteracy + HS.Grad + Frost + Area, data=statedata, test="F")
# we add Frost
model2 = lm(Murder ~ Life.Exp + Frost, data=statedata)
add1(model2, scope= ~ . + Population + Income + Illiteracy + HS.Grad +  Area, data=statedata, test="F")
# we add Area
model3 = lm(Murder ~ Life.Exp + Frost + Area, data=statedata)
add1(model3, scope= ~ . + Population + Income + Illiteracy + HS.Grad, data=statedata, test="F")
# we add Population
model4 = lm(Murder ~ Life.Exp + Frost + Area + Population, data=statedata)
add1(model4, scope= ~ . + Income + Illiteracy + HS.Grad, data=statedata, test="F")
# we add Illiteracy
model5 = lm(Murder ~ Life.Exp + Frost + Area + Population + Illiteracy, data=statedata)
add1(model5, scope= ~ . + Income + HS.Grad, data=statedata, test="F")
# Our final model keeps Life.Exp, Frost, Area, Population, Illiteracy

model5
```

Question 4. c)
```{r q4c}
fullmodel = lm(Murder ~ Population + Income + Illiteracy + Life.Exp + HS.Grad + Frost + Area, data=statedata)
model = step(fullmodel, scope= ~ . + Population + Income + Illiteracy + Life.Exp + HS.Grad + Frost + Area, data=statedata)
# Our final model keeps Life.Exp, Area, Illiteracy, Population, Frost at a significance level of alpha=0.05

model
```

Question 4. e)
```{r q4e}
plot(model)
```
Compare this to another final model that also used a log transformation on Population.
```{r other}
statedata$Population = log(statedata$Population)

fullmodel = lm(Murder ~ Population + Income + Illiteracy + Life.Exp + HS.Grad + Frost + Area, data=statedata)
model = step(fullmodel, scope= ~ . + Population + Income + Illiteracy + Life.Exp + HS.Grad + Frost + Area, data=statedata)

plot(model)
```
Not taking a transformation is arguably better even though it looks like it should have.
Residuals vs Fitted are not as spread, QQ-Plot suggests that the tails follow another distribution, Scale-Location seems to have a negative quadratic trend and the Residuals vs Leverage has points with much larger leverage compared to the previous final model.

Question 5. b)
```{r q5b}
Xscaled = scale(X[,-1]) # No intercept parameter (Piazza)
yscaled = scale(y, scale=FALSE) # Only centering, no scale (Piazza)
r = dim(t(Xscaled)%*%Xscaled)
lambda = diag(0.5, r)
b = solve(t(Xscaled)%*%Xscaled + lambda, t(Xscaled)%*%yscaled)

b
```

Question 5. c)
```{r q5c}
library(matrixcalc)
aic = c()
lambdas = seq(0, 0.5, by=0.01)

for (i in lambdas){
  lambda = diag(i, r)
  
  ridgeb = solve(t(Xscaled)%*%Xscaled + lambda, t(Xscaled)%*%yscaled)
  SSRes = t(yscaled-Xscaled%*%ridgeb)%*%(yscaled-Xscaled%*%ridgeb)
  
  H = Xscaled%*%solve(t(Xscaled)%*%Xscaled + lambda)%*%t(Xscaled)
  
  aic = c(aic, n*log(SSRes/n) + 2*matrix.trace(H))
}

plot(lambdas, aic, col='orange', type='l')
lambda_aic = lambdas[which.min(aic)]

lambda_aic
```


















































